Re-Allocation of Resources during Releases \titlenote{Copyright note}

### 2013-12-03 paper1/paper1.tex
updating paper. started writing conclusion

\title{Re-Allocation of Resources during Releases \titlenote{Copyright note}}
\numberofauthors{2}
\author{
% 1st. author
\alignauthor
Md Tajmilur Rahman\\
       \affaddr{Concordia University}\\
       \affaddr{Montreal, QC H3G 1M8}\\
       \affaddr{438-932-2288, +1}\\
       \email{mdt\_rahm@encs.concordia.ca}
% 2nd. author
\alignauthor
Peter C. Rigby\\
       \affaddr{Concordia University}\\
       \affaddr{Montreal, QC H3G 1M8}\\
       \affaddr{514-848-2424, +1}\\
       \email{peter.rigby@concordia.ca}
}
\date{20 Oct 2013}
\maketitle
\begin{abstract}
This section will be written at the end.
\end{abstract}
\category{K.6.3}{Software Management}{Software Development, Software Resource Management, Resource Reallocation}
\terms{Experiment, Human Factors, Resource Management, Reallocation}
\keywords{Resource Reallocation, Code Ownership, Defect Density, Software Releases} % NOT required for Proceedings
\section{Introduction}
Software projects are notorious for going over budget and schedule. Rush periods are often get seen before a major release that turn the developers into dinosaurs as Frederick Brooks likens in his benchmark study "The Mythical Man Month" \cite{brooks_mythical}. This "Rush To Release (RTR)" can be prompted either by external forces such as decisions by management to include new features in the release or to release earlier to beat a competitor. Alternatively, the rush may simply be due to inappropriate or unrealistic scheduling. Whatever the reason is it is an obvious. Regardless of the causes, the rush to release stresses developers and often requires developers to work on unusual, high priority or critical areas of the system. In this paper we study how RTR may effect project organization and introduce technical debt. We want to propose a method to observe, analyze and summarize the results of revisions found around releases. We intend to infer behavior of the developers related to the process around release time. The key research questions that we expect to answer with our methodology are described in section 1.1.

@@

We observe the developers' working areas to understand the allocation of the resources within an open source software development project to understand the reallocation or distribution takes place in a software development process which may cause disruptive events. We attempt to identify the project's different release times and calculate the difference between different development periods within a releases cycle to identify which new areas are being worked on and what is the behavior of the developers between normal development period and the period when RTR is prompted. The historical data from git that we are working on for this purpose will help us to extract a lot of information like calculating the developers' working areas and time-frame of each release as well as different periods in a stable release period. This information will help us to identify the criteria of the resources, their roles and behaviors, code ownerships of the domains they are working in.

@@

\subsection{Research Questions}
We have set our research questions focusing the key properties of software development process, allocation of developers throughout the process, especially for open source software development. First couple of research questions are to understand the basic structure and strategies of the development and release and merging process practiced by Linux Kernel development team. Rest of the questions are to understand the distribution of developers' working areas and the behavior of developers in different segments of a release period.

\renewcommand{\labelenumi}{Q\theenumi:}
\begin{enumerate}
\item What is the release process used by the project? \newline
We will be giving a qualitative description of Linux Kernel development in brief to give anser to this question.
\item How many developers are allocated troughout different segments of a release period? What significant difference can be observed between the segments of an entire release period? \newline
Answer to this question will give us a big picture of Linux Kernel development community as well as the contrubution of developers in different periods within a stable release period. We will calculate Jaccard Distance to differentiate the behavior of developers during release period and development period.
\item Do developers work on different areas of the system around the time of release? \newline
We want to see if developers are looking at different types of files in the code-base around the time of release which is being considered as the release period or rc period. We are expecting this would be the RTR in a release cycle which may introduce technical debts.
\item Do developers work in others' code in a high proportion during the rush period?\newline
Developers work on various files during release period. We will try to find out what is the percentage of dealing with files that developers own or they are most familier to. This will give us a sense of the nature of work during release time.
\item How long a commit takes to get pushed into the main kernel? Does it differ in different periods of release?\newline
In distributed development environment developers work in different branches they commit their codes to the branches they are associated to. Developers in that case do not directly push their changes to the main branch. We would like to see how much time it takes for a commit to the local branch to get pushed to the main branch. We also should understand how long it takes for regular development commits and how long it takes for the commits made around the release period.
\item Are there certain areas of the system that receive increased attention (i.e.\ do developers focus on a smaller set of files around releases)? \newline
To answer this question we will look into the files getting churned extremely high in number. We will try to figure out the difference between those highly churned or highly focused sets of files in different releases to understand if they require contantly high focus or they are only for one particular release.
\end{enumerate}

@@

\subsection{Finding Commits for Releases}
We have to sort out and accumulate all the commits together those are belonging to any of the releases that we have. We have all 400441 commits in our hand but though we can't plot them into a single time domain to assign different set of commits to different release periods because development goes in parallel on different branches in git so commits also go in parallel. If there are multiple branches, surely there are multiple merges and the commits prior to a crisscross merge are all ancestors of the heads of all subsequent branches \cite{bird_git}. Figure 1 shows how a git DAG looks like having so many crisscross merges in a busy branching snap-shot of the git DAG of Linux Kernel development.
\begin{figure}
\begin{center}
\includegraphics[height=2.5in,width=2.5in]{gitdag.png}
\caption{\small \sl Branching in git DAG for linux kernel development.}
\end{center}
\end{figure}
So it is possible to make a commit months ago from today while several releases may have already been made in the mean time but that commit can belong to any future release. Commits that are being made to a branch may get merged with other branch or the head branch during the release period. This is why we had to track all the way down the trees for each of the 340 releases in the git DAG to find out which commit contributes to which release. We could do this using Git DAG Traversing (GDT) algorithm as shown in Algorithm 1. After applying GDT we collected 381152 out of 400441 commits those are well organized in the git DAG for Linux Kernel development. As we mentioned earlier that we have last couple of release candidates for ``linuxv2.6.12'' and first couple of release candidates for ``linuxv3.12''. We are choping out them because we don't have complete release commit data for version 2.6.12-rc as well as 3.12-rc releases. Out of 400441 commits 19289 are belonging to those releases. So finally we have 381152 commits for 340 releases in our database to progress our work on.

@@

\begin{table}[ht]
\caption{Code Ownership in RP}  % title of Table
\centering 						% used for centering table
\begin{tabular}{c c c c}				% centered columns (4 columns)
\hline\hline						%inserts double horizontal lines
% table heading
Author 				& Linuxv		& Path				& $\omega$ (\%) \\ [0.5ex]
\hline 							% single horizontal line
% inserting body of the table
Thomas Gleixner		& 2.6.24		& .../numa\_64.h	& 0.0000\\
Jeff Kirsher			& 3.1		& .../ethtool.c		& 0.0000\\
Sathya Perla			& 2.6.29		& .../hwlib.h		& 100.00\\
Roel Kluin			& 2.6.24		& .../innovator.h 	& 100.00\\
David S. Miller		& 2.6.24		& .../visasm.h 		& 100.00\\
Stephen Hemminger	& 2.6.24		& .../qla3xxx.c	 	& 12.083\\
Randy Dunlap			& 2.6.24		& ...\_core.c 		& 100.00\\
[1ex]							% adds vertical space
\hline 							% inserts single line
\end{tabular}
\label{table:nonlin} 				% is used to refer this table in the text
\end{table}
The example given above is calculated for finding ownerships of developers worked for different files in different releases. Ownership is calculated for all the developers in release periods as well as in merge period. In addition we have calculated which file is owned by which developer and also the percentage of working with owned files during merge meriod and release period seperately. These extractions will help us drawing the results from our findings to answer our 3rd and 4th research questions.

@@

In the early 1990's the Linux kernel development was not so busy affair \cite{linux_kernel}. There were very small number of users and developers involved at that time. Developers use a loosely time-based release process. Linux publishes it's major kernel release in every two or three months. If we look at figure 2 the upper Line represents how the stable releases are going on since 2005. From the figure we can see the average duration of each stable release is 62.47 although ``linuxv2.6.24'' was significantly hight as exception, it took 92.8 days. The bottom line which is much steady than the upper one is representing the durations of merge period in each stable release. If we put the releases in a time domain we can see a picture shown in figure 3 where we see that the stable releases are all going on maintaining almost a similar frequency and similar pattern. The highest picks are the merge periods then the smaller picks between the highest ones are representing the rc releases. For more clear view we have plotted only 100 releases of all 340 releases including rc candidates.

@@

\subsection{Difference between release segments}
\textit{How many developers are allocated troughout different segments of a release period? What significant differ-ence can be observed between the segments of an entire release period?}

In order to see how many developers contribute to the new featcher developments and how many contribute in merging and fixing, review and accepting codes from the community during the merge period, we observed the developers' area in Merge Period and developers' area in Release Period. Out of all 381152 commits belonging to all 340 releases 292926 commits were made during the time of merge period which is 76.85\% (more than 2/3) of the total. These commits were made by 14621 individual developers. We now keeping only the records belong to release 2.6.13 to 3.11 because we have complete information for thse releases, so we had to remove the commit records belong to linuxv2.6.12 and linuxv3.12 and later. 11209 developers work in merge period and 7625 developers work around the release period.

@@

Once we found different releases and the release periods as well as we calculate the developers' areas and code ownerships for developers we also would like to know what is the difference being created by making changes between the two main segments within an entire release development cycle (i.e.\ Merge Period MP and Release Period RP). To observe this we want to calculate the difference using Jaccard Similarity Coefficient \cite{jaccard_alpine} which is represented by equation 3.
\begin{equation} J(A, B) =\frac{|A \cap B|}{|A \cup B|} \end{equation}
The Jaccard Distance can be obtained by subtracting the Jaccard Coefficient from 1.
\begin{equation} d_j(A, B) = 1 - J(A, B) = \frac{|A \cup B|-|A \cap B|}{|A \cup B|} \end{equation}
Here A and B are two sets. For calculating Jaccard Distance between two releases we are considering the files that have been worked for, in a particular release. If set A is the set of files worked in MP and files worked for in another release is set B then In figure 6 the Jaccard distances between MP and RP for each release can be represented. The average jaccard distance between merge period and release period for all the release cycles is 0.85 and we see the jaccard distance is also going upward by releases except the only exception at release 2.6.34.

@@

\subsection{Developers' contributions around the time of release}
\textit{Do developers work on different areas of the system around the time of release?}\newline
\textit{Do developers work in others' code in a high proportion during the rush period?}

@@

\subsection{Time for commits to go live}
\textit{How long a commit takes to get pushed into the main kernel? Does it differ in different periods of release?}

@@

\subsection{Developers' focus on code-base around the time of release}
\textit{Are there certain areas of the system that receive increased attention (i.e.\ do developers focus on a smaller set of files around releases)?}

@@

So, we can say that there are certain areas of the system that takes certainly increased attention also there are few developers who deals with extremely large number of files at the time of release. Some of them work less than 40\% files that they own and some developers work around 70\% of files that they own. Now we would like to see if these areas having increased attention get the similar attention in other releases. In this occassion we would like find the Jaccard similarity index for these areas in consicutive releases. First we collected the sets of files those are having highly increasing focus during the release periods of different release cycles. Then we calculated the union and intersection operation between two consicutive sets as well as randomly. We applied equation 2 to find  out the similarity among the sets and we see there is no similarity between those sets except for couple of pairs of releases like linuxv2.6.14~linuxv2.6.15,  linuxv2.6.29~linuxv2.6.30, linuxv2.6.30~linuxv2.6.31 and linuxv2.6.34~linuxv2.6.35 where the similarity index is too low to say similar; they are 0.038, 0.066, 0.025 and 0.181 respectively as shown in figure 14. From this we understand that developers working for different and various features, developing new staffs for linux kernel working in many different files and merging them into the main kernel during merge window. In release period small number of developers who are allowed to work in release period are reviewing those nely arrived codes and doing quick fixes and enhancements, some of them may require very high attention and more work to do. In the next release some other different new developments arrive in merge window and thus in release period developer do their candidate releases focusing on those newly arrived code files which are not related to the code files those had to get high focus in the previous release.

@@

\section{Conclusions}
This section will be written at the end.
%\end{document}  % This is where a 'short' article might terminate

>>>>

\title{Re-Allocation of Resources during Releases \titlenote{Copyright note}}
\numberofauthors{2}
\author{
% 1st. author
\alignauthor
Md Tajmilur Rahman\\
       \affaddr{Concordia University}\\
       \affaddr{Montreal, QC H3G 1M8}\\
       \affaddr{438-932-2288, +1}\\
       \email{mdt\_rahm@encs.concordia.ca}
% 2nd. author
\alignauthor
Peter C. Rigby\\
       \affaddr{Concordia University}\\
       \affaddr{Montreal, QC H3G 1M8}\\
       \affaddr{514-848-2424, +1}\\
       \email{peter.rigby@concordia.ca}
}
\date{20 Oct 2013}
\maketitle
\begin{abstract}
This section will be written at the end.
\end{abstract}
\category{K.6.3}{Software Management}{Software Development, Software Resource Management, Resource Reallocation}
\terms{Experiment, Human Factors, Resource Management, Reallocation}
\keywords{Resource Reallocation, Code Ownership, Defect Density, Software Releases} % NOT required for Proceedings
\section{Introduction}
Software projects are notorious for going over budget and schedule. Rush periods are often get seen before a major release that turn the developers into dinosaurs as Frederick Brooks likens in his benchmark study "The Mythical Man Month" \cite{brooks_mythical}. This "Rush To Release (RTR)" can be prompted either by external forces such as decisions by management to include new features in the release or to release earlier to beat a competitor. Alternatively, the rush may simply be due to inappropriate or unrealistic scheduling. Whatever the reason is it is an obvious. Regardless of the causes, the rush to release stresses developers and often requires developers to work on unusual, high priority or critical areas of the system. In this paper we study how RTR may effect project organization and introduce technical debt. We want to propose a method to observe, analyze and summarize the results found around different parts of a release cycle. We intend to infer behavior of the developers related to the process around release time. The key research questions that we expect to answer with our methodology are described in section 1.1.

@@

We observe the developers' working areas to understand the allocation of the resources within an open source software development project to understand the reallocation or distribution takes place in a software development process which may cause disruptive events. We attempt to identify the project's different release times and calculate the difference between different development periods within a releases cycle to identify which new areas are being worked on and what is the behavior of the developers between normal development period and the period when RTR is prompted. The historical data from git that we are working on for this purpose will help us to extract a lot of information like calculating the developers' working areas and time-frame of each release as well as different parts in a stable release cycle. This information will help us to identify the criteria of the resources, their roles and behaviors, code ownerships of the domains they are working in, density of work, nature of dealing with files around the time of release etc.

@@

\subsection{Research Questions}
We have set our research questions focusing the key properties of software development process, allocation of developers throughout the process, especially for open source software development. First couple of research questions are to understand the basic structure and strategies of the development and release and merging process practiced by Linux Kernel development team. Rest of the questions are to understand the distribution of developers' working areas and the behavior of developers in different segments of a release cycle.

\renewcommand{\labelenumi}{RQ\theenumi:}
\begin{enumerate}
\item What is the release process used by the project? \newline
We will be giving a qualitative description of Linux Kernel development in brief to give anser to this question.
\item How many developers are allocated troughout different segments of a release cycle? What significant difference can be observed between the segments of a release cycle? \newline
Answer to this question will give us a big picture of Linux Kernel development community as well as the contrubution of developers in different parts within a stable release.
\item Do developers work on different areas of the system around the time of release? \newline
We want to see if developers are looking at different types of files in the code-base around the time of release which is being considered as the release period or rc period. We are going to introduce an algorithm to traverse through the git DAG to find out commits associated to a particular release. This can be useful for any other data collected from distributed version control system like git. We are expecting this would be the RTR in a release cycle which may introduce technical debts.
\item Do developers work in others' code in a high proportion during the rush period?\newline
Developers work on various files during release period. We will try to find out what is the percentage of dealing with files that developers own or they are most familier to. This will give us a sense of the nature of work during release time.
\item How long a commit takes to get pushed into the main kernel? Does it differ in different periods of release?\newline
In distributed development environment developers work in different branches. They commit their codes to the branches they are associated to. While core development is going on developers  do not directly push their changes to the main branch. We would like to see how much time it takes for a commit to the local branch to get pushed to the main branch. We also should understand how long it takes for regular development commits and how long it takes for the commits made around the release period.
\item Are there certain areas of the system that receive increased attention (i.e.\ do developers focus on a smaller set of files around releases)? \newline
To answer this question we will look into the files getting churned extremely high in number. We will try to figure out the difference between those highly churned or highly focused sets of files in different releases to understand if they require contantly high focus or they are only for one particular release.
\end{enumerate}

@@

\subsection{Finding Commits for Releases}
We have to sort out and accumulate all the commits together those are belonging to any of the releases. We have all 400441 commits in our hand but though we can't plot them into a single time domain to assign different set of commits to different release periods because development goes in parallel on different branches in git, so commits also go in parallel. If there are multiple branches, surely there are multiple merges and the commits prior to a crisscross merge are all ancestors of the heads of all subsequent branches \cite{bird_git}. Figure 1 shows how a git DAG looks like having so many crisscross merges in a busy branching snap-shot of the git DAG of Linux Kernel development.
\begin{figure}
\begin{center}
\includegraphics[height=2.5in,width=2.5in]{gitdag.png}
\caption{\small \sl Branching in git DAG for linux kernel development.}
\end{center}
\end{figure}
So it is possible to make a commit months ago from today while several releases may have already been made in the mean time but that commit can belong to any future release. Commits that are being made to a branch may get merged with other branch or the head branch during the release period. This is why we had to track all the way down the trees for each of the 340 releases in the git DAG to find out which commit contributes to which release. We could do this using Git DAG Traversing (GDT) algorithm as shown in Algorithm 1. After applying GDT we collected 381152 out of 400441 commits those are well organized in the git DAG of Linux Kernel development. As we mentioned earlier that we have last couple of release candidates for ``linuxv2.6.12'' and first couple of release candidates for ``linuxv3.12''. We are choping them out because we don't have complete commit records for version 2.6.12-rc releases as well as 3.12-rc releases. Out of 400441 commits 19289 are belonging to those releases. Therefore, finally we have 381152 commits for 331 releases in our database to progress our work on.

@@

\begin{table}[ht]
\caption{Code Ownership in Release Period}  % title of Table
\centering 						% used for centering table
\begin{tabular}{c c c c}				% centered columns (4 columns)
\hline\hline						%inserts double horizontal lines
% table heading
Author 				& Linuxv		& Path				& $\omega$ (\%) \\ [0.5ex]
\hline 							% single horizontal line
% inserting body of the table
Thomas Gleixner		& 2.6.24		& .../numa\_64.h	& 0.0000\\
Jeff Kirsher			& 3.1		& .../ethtool.c		& 0.0000\\
Sathya Perla			& 2.6.29		& .../hwlib.h		& 100.00\\
Roel Kluin			& 2.6.24		& .../innovator.h 	& 100.00\\
David S. Miller		& 2.6.24		& .../visasm.h 		& 100.00\\
Stephen Hemminger	& 2.6.24		& .../qla3xxx.c	 	& 12.083\\
Randy Dunlap			& 2.6.24		& ...\_core.c 		& 100.00\\
[1ex]							% adds vertical space
\hline 							% inserts single line
\end{tabular}
\label{table:nonlin} 				% is used to refer this table in the text
\end{table}
The example given above is calculated for finding ownerships of developers worked for different files in different releases. Ownership is calculated for all the developers in release periods as well as in merge period. In addition we have calculated which file is owned by which developer and also the percentage of working with owned files during merge meriod and release period seperately. These extractions will help us drawing the results from our findings to answer our 3rd and 4th research questions.

@@

In the early 1990's the Linux kernel development was not so busy affair \cite{linux_kernel}. There were very small number of users and developers involved at that time. Developers use a loosely time-based release process. Linux publishes it's major kernel release in every two or three months. If we look at figure 2 the upper Line represents how the stable releases are going on since 2005. From the figure we can see the average duration of each stable release is 62.47 although ``linuxv2.6.24'' was significantly hight as exception, it took 92.8 days. The bottom line which is much steady than the upper one is representing the durations of merge period in each stable release. If we put the releases in a time domain we can see a picture shown in figure 3 where we see that the stable releases are all going on maintaining almost a similar frequency and similar pattern. The highest picks are the merge periods then the smaller picks between the highest ones are representing the rc releases. For more clear view we have plotted only 100 releases of all 331 releases including rc candidates.

@@

\subsection{Difference Between Release Segments}
\textit{How many developers are allocated troughout different segments of a release period? What significant differ-ence can be observed between the segments of an entire release period?}

In order to see how many developers contribute to the new featcher developments and how many contribute in merging and fixing, review and accepting codes from the community during the merge period, we observed the developers' area in Merge Period and developers' area in Release Period. Out of all 381152 commits belonging to all 331 releases 292926 commits were made during the time of merge period which is 76.85\% (more than 2/3) of the total. These commits were made by 14621 individual developers. We now keeping only the records belong to release 2.6.13 to 3.11 because we have complete information for thse releases, so we had to remove the commit records belong to linuxv2.6.12 and linuxv3.12 and later. 11209 developers work in merge period and 7625 developers work around the release period.

@@

Once we found different releases and the release periods as well as we calculate the developers' areas and code ownerships for developers we also would like to know what is the difference being created by making changes between the two main segments within an entire release development cycle (i.e.\ Merge Period MP and Release Period RP). To observe this we want to calculate the difference using Jaccard Similarity Coefficient \cite{jaccard_alpine} which is represented by equation 3.
\begin{equation} J(A, B) =\frac{|A \cap B|}{|A \cup B|} \end{equation}
The Jaccard Distance can be obtained by subtracting the Jaccard Coefficient from 1.
\begin{equation} d_j(A, B) = 1 - J(A, B) = \frac{|A \cup B|-|A \cap B|}{|A \cup B|} \end{equation}
Here A and B are two sets. For calculating Jaccard Distance between two releases we are considering the files that have been worked for, in a particular release. If set A is the set of files worked in MP and files worked for in another release is set B then In figure 6 the Jaccard distances between Merge Period and Release Period for each release can be represented. The average jaccard distance between merge period and release period for all the release cycles is 0.85 and we see the jaccard distance is also going upward by releases except the only exception at release 2.6.34.

@@

\subsection{Developers' Contributions Around the Time of Release}
\textit{Do developers work on different areas of the system around the time of release?}\newline
\textit{Do developers work in others' code in a high proportion during the rush period?}

@@

\subsection{Time for Commits to Go Live}
\textit{How long a commit takes to get pushed into the main kernel? Does it differ in different periods of release?}

@@

\subsection{Developers' Focus on Code-Base Around the Time of Release}
\textit{Are there certain areas of the system that receive increased attention (i.e.\ do developers focus on a smaller set of files around releases)?}

@@

So, we can say that there are certain areas of the system that takes certainly increased attention also there are few developers who deal with extremely large number of files at the time of release. Some of them work less than 40\% files that they own and some developers work around 70\% of files that they own. Now we would like to see if these areas having increased attention get the similar attention in other releases. This finding also supports our fourth research question \textit{RQ4} strongly. In this occassion we would like find the Jaccard similarity index for these areas in consicutive releases. First we collected the sets of files those are having highly increasing focus during the release periods of different release cycles. Then we calculated the union and intersection operation between two consicutive sets as well as randomly. We applied equation 2 to find  out the similarity among the sets and we see there is no similarity between those sets except for couple of pairs of releases like linuxv2.6.14~linuxv2.6.15,  linuxv2.6.29~linuxv2.6.30, linuxv2.6.30~linuxv2.6.31 and linuxv2.6.34~linuxv2.6.35 where the similarity index is too low to say similar; they are 0.038, 0.066, 0.025 and 0.181 respectively as shown in figure 14. From this we understand that developers working for different and various features, developing new staffs for linux kernel working in many different files and merging them into the main kernel during merge window. In release period small number of developers who are allowed to work in release period are reviewing those nely arrived codes and doing quick fixes and enhancements, some of them may require very high attention and more work to do. In the next release some other different new developments arrive in merge window and thus in release period developer do their candidate releases focusing on those newly arrived code files which are not related to the code files those had to get high focus in the previous release.

@@

\section{Conclusions}
The results have been reported with respect to our research questions in this case study. We reported following items specifically:
\renewcommand{\labelenumi}{\theenumi:}
\begin{enumerate}
\item{The basic structure and the development process of Linux Kernel}
\item{The distribution of developers in two different segments in a complete release cycle}
\item{Role of developers during merge period and the significance of release period}
\item{Developers' focus around the time of release, percentage of working with native code during merge period and release period.}
\end{enumerate}
%\end{document}  % This is where a 'short' article might terminate
